%!TEX root = vaisagh_thesis.tex

\begin{acknowledgements}

I would like to express my sincere gratitude to my supervisor, \textbf{Dr. Bo An} for his incredible understanding, help and support in a very difficult situation. I would also like to thank my co-supervisor and mentor, \textbf{Dr. Michael Lees} for giving me the opportunity and the inspiration to pursue my dreams. Right from the days of co-supervising my Final Year Project, he has always been a reliable and uncompromising critic who's been there to help, advise and discuss about everything ranging from my research to teaching to technology and life in general. The weekly team meetings over coffee and, more recently, over hangouts, and his emphasis on encouraging everyone in his team to discuss their projects with each other not only helped me develop a lot of my ideas but made PhD life a lot more interesting when things got boring; and bearable when the going got tough.

\textbf{Prof. Peter Sloot} has my sincerest gratitude for taking time out of his busy schedule to give me valuable feedback, advise and support for my research. I would also like to thank \textbf{Dr. Cheong Siew Ann} for giving me the chance to work with and learn from him. A special thanks goes out to \textbf{Prof. Cai Wentong} for introducing me to Crowd Simulation and giving me the chance to be a part of several interesting projects over the past several years.

I'm grateful to \textbf{Dr. Heiko Aydt} and \textbf{Dr. Hu Nan} for not only letting me be a part of some of the really interesting work they were doing but, more importantly, for sharing their experience in working on a PhD at NTU which helped me avoid and handle most of the major and minor hurdles that I faced as a PhD Candidate. I must also acknowledge \textbf{Debraj Roy} and \textbf{Nasri Bin Othman} for being sounding boards for my ideas and friends in an environment where I've sometimes felt lost. My thesis would not have been complete without the invaluable help of \textbf{Cheong Eu Lee} in making sense of the numbers.

I would also like to express my sincere gratitude to the School of Computer Engineering, NTU and the Parallel and Distributed Computing Center for providing me with the opportunity, the funds and, above all, the excellent facilities to pursue a PhD. A special thanks is also due to \textbf{Irene} (Ng-Goh Siew Lai), \textbf{Kenneth} (Poliran Kenneth Caballes) and \textbf{Saen} (Lok Chun Saen) for making working at PDCC so easy and convenient.

Finally, I would like to thank my family for their unflinching support and belief in me; and \textbf{Aparna} for listening to my constant complaints and rants, and always being there to cheer me up.

Thank you
\\
\\
\hspace{0in}
Vaisagh Viswanathan T\\
\today\\
{Nanyang Technological University, Singapore}\\
\end{acknowledgements}